Analyzing changes in Canadian grocery prices

### 2024-11-14 paper/paper.qmd
Updated intro, measurement, and abstract

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

@@

## Outcome variables

## Predictor variables

@@

```{r}
#| label: fig-avg-curr-price
#| fig-cap: Average Current Price by Vendor. This bar plot shows the average price of products sold by each grocery vendor in Canadian dollars (CAD). Variations in pricing suggest differences in market positioning or pricing strategies among vendors, with some retailers consistently offering higher or lower prices across product categories.
#| echo: false
#| warning: false
#| message: false

#### Read data ####
# Load the cleaned dataset
analysis_data <- read_csv("C:/Users/eliza/Downloads/hammer-3-compressed/cleaned_price_changes.csv", show_col_types = FALSE)

# Generate plot for Average Current Price by Vendor
vendor_price_plot <- analysis_data %>%
  group_by(vendor_name) %>%
  summarize(avg_price_current = mean(price_current, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(vendor_name, avg_price_current), y = avg_price_current)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(
    title = "Average Current Price by Vendor",
    x = "Vendor",
    y = "Average Current Price (CAD)"
  ) +
  theme_minimal()

# Display the plot
vendor_price_plot

@@

@@

```{r}
#| label: fig-avg-price-diff
#| fig-cap: Average Price Difference by Vendor. This bar plot shows the average price change per vendor, highlighting retailers with the most significant pricing adjustments. Positive differences suggest recent price increases, while negative differences indicate reductions, reflecting possible discounting or competitive pricing strategies.
#| echo: false
#| warning: false
#| message: false

# Average price difference by vendor
price_diff_plot <- analysis_data %>%
  group_by(vendor_name) %>%
  summarize(avg_price_difference = mean(price_difference, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(vendor_name, avg_price_difference), y = avg_price_difference)) +
  geom_bar(stat = "identity", fill = "coral") +
  coord_flip() +
  labs(
    title = "Average Price Difference by Vendor",
    x = "Vendor",
    y = "Average Price Difference (CAD)"
  ) +
  theme_minimal()
price_diff_plot

@@

@@

```{r}
#| label: fig-past-curr
#| fig-cap: Relationship Between Previous and Current Prices. This scatter plot with a linear regression line shows the correlation between previous and current prices. A positive relationship would indicate that higher previous prices predict higher current prices, implying potential price stability or gradual price changes in the grocery market.
#| echo: false
#| warning: false
#| message: false

# Scatter plot of previous price vs. current price
price_relationship_plot <- ggplot(analysis_data, aes(x = price_old, y = price_current)) +
  geom_point(alpha = 0.4, color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Relationship Between Previous and Current Prices",
    x = "Previous Price (CAD)",
    y = "Current Price (CAD)"
  ) +
  theme_minimal()
price_relationship_plot

@@

```{r}
#| label: fig-dist-price
#| fig-cap: Relationship Between Previous and Current Prices. This scatter plot with a linear regression line shows the correlation between previous and current prices. A positive relationship would indicate that higher previous prices predict higher current prices, implying potential price stability or gradual price changes in the grocery market.
#| echo: false
#| warning: false
#| message: false

# Calculate mean and standard deviation of price_difference
mean_diff <- mean(analysis_data$price_difference, na.rm = TRUE)
sd_diff <- sd(analysis_data$price_difference, na.rm = TRUE)

# Filter out extreme values beyond 3 standard deviations
filtered_data <- analysis_data %>%
  filter(price_difference >= (mean_diff - 3 * sd_diff),
         price_difference <= (mean_diff + 3 * sd_diff))

# Plot histogram with filtered data
price_difference_hist <- ggplot(filtered_data, aes(x = price_difference)) +
  geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
  labs(
    title = "Distribution of Price Differences (Filtered)",
    x = "Price Difference (CAD)",
    y = "Frequency"
  ) +
  theme_minimal()
price_difference_hist

@@

@@

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

>>>>

---
title: "Analyzing changes in Canadian grocery prices"
# subtitle: "My subtitle if needed"
author: 
  - Abdullah Motasim
  - Elizabeth Luong
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "This paper investigates grocery pricing trends across Canada using SQL analysis on the 'Project Hammer' dataset, which contains detailed vendor-specific price data. Focusing on statistical relationships and potential pricing biases, we explore factors driving price variations across different regions and vendors. A Bayesian regression model is applied to understand correlations, accounting for limitations and highlighting areas where causation cannot be assumed. Findings reveal competitive dynamics within Canadian grocery markets, implications for consumers, and recommendations for future research into market and pricing behavior across vendors."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

The rising cost of groceries in Canada has drawn significant public attention, impacting consumer budgets and prompting regulatory interest. While inflation and supply disruptions are often cited as contributors, the specifics of how major Canadian grocery vendors set and adjust prices remain underexplored. This study addresses this gap by analyzing data from Project Hammer, a dataset that tracks grocery prices across vendors including Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods. Through examining these data, we aim to uncover factors driving grocery price dynamics and assess the degree to which vendor-specific pricing may reflect competitive or non-competitive practices.

Using a Bayesian regression model, our analysis identifies prior prices, vendor identity, and recent price changes as significant determinants of current grocery prices. The consistency in historical pricing indicates relative stability across product categories, while unique vendor patterns suggest differing market strategies, likely shaped by operational costs and customer demographics.

Our findings have practical implications for both consumers and policymakers: consumers can use vendor-specific insights to make informed purchasing decisions, while policymakers can leverage this analysis to support regulatory strategies that promote competition and price transparency. The paper is structured as follows: Section 2 reviews the dataset, Section 3 describes the Bayesian model, Section 4 presents the results, and Section 5 discusses broader implications and future research directions.

@@

Each entry in the dataset represents a captured moment in time where prices, products, and vendor information were recorded. The measurement process involved regular data collection from the online pricing systems of the following prominent grocery chains: Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods. Each entry captures variables such as:  product name, vendor name, and both current and previous prices. The Project Hammer initiative leveraged web scraping technology to ensure a consistent and standardized entry format for all retailers, documenting prices at regular intervals. This approach ensures that price points reflect real-time data rather than historical estimates or annual averages, providing a close approximation of consumer experiences. The data set records the price of each grocery item in Canadian dollars, including items from broad categories such as fresh produce, dairy, pantry goods, and household essentials, to present a holistic view of grocery costs.

<!-- Each row entry in Project Hammer represents a unique product listing per vendor, featuring five main columns: product_name, vendor_name, price_current, price_old, and price_difference. The price_difference column is particularly useful for observing temporal price changes, while the price_old variable provides context for price continuity or volatility. The structured approach used in capturing this data allows us to transform general observations of grocery pricing trends into quantifiable entries that can be statistically analyzed to reveal underlying patterns in pricing strategies across different vendors and products. -->

Essentially, the real world phenomenon we observed was the website of each vendor with a listing of their products for that week, we turned this phenomenon into an entry within out data set with the use of a screen-scrape of the website UI. This means a HTTP request to load the page was sent to the desired vendors website on a specific day, than the returned HTML was parsed to extract specific content such as text, images, prices, etc. These captured features were than utilized to fill out the corresponding columns within the data set.

@@

```{r}
#| label: fig1
#| fig-cap: Each boxplot shows the distribution of numeric reliability grades for the top 5 pollsters, with individual poll grades represented by jittered points.
#| echo: false
#| warning: false
#| message: false

#### Read data ####
# Load the cleaned dataset
# analysis_data <- read_csv("C:/Users/eliza/Downloads/hammer-3-compressed/cleaned_price_changes.csv", show_col_types = FALSE)
# 
# # Generate plot for Average Current Price by Vendor
# vendor_price_plot <- analysis_data %>%
#   group_by(vendor_name) %>%
#   summarize(avg_price_current = mean(price_current, na.rm = TRUE)) %>%
#   ggplot(aes(x = reorder(vendor_name, avg_price_current), y = avg_price_current)) +
#   geom_bar(stat = "identity", fill = "skyblue") +
#   coord_flip() +
#   labs(
#     title = "Average Current Price by Vendor",
#     x = "Vendor",
#     y = "Average Current Price (CAD)"
#   ) +
#   theme_minimal()
# 
# # Display the plot
# vendor_price_plot

@@

Figure 1: Average Current Price by Vendor. This bar plot shows the average price of products sold by each grocery vendor in Canadian dollars (CAD). Variations in pricing suggest differences in market positioning or pricing strategies among vendors, with some retailers consistently offering higher or lower prices across product categories.

@@

```{r}
#| label: fig2
#| fig-cap: Each boxplot shows the distribution of numeric reliability grades for the top 5 pollsters, with individual poll grades represented by jittered points.
#| echo: false
#| warning: false
#| message: false

# Average price difference by vendor
# price_diff_plot <- analysis_data %>%
#   group_by(vendor_name) %>%
#   summarize(avg_price_difference = mean(price_difference, na.rm = TRUE)) %>%
#   ggplot(aes(x = reorder(vendor_name, avg_price_difference), y = avg_price_difference)) +
#   geom_bar(stat = "identity", fill = "coral") +
#   coord_flip() +
#   labs(
#     title = "Average Price Difference by Vendor",
#     x = "Vendor",
#     y = "Average Price Difference (CAD)"
#   ) +
#   theme_minimal()
# price_diff_plot

@@

Figure 2: Average Price Difference by Vendor. This bar plot shows the average price change per vendor, highlighting retailers with the most significant pricing adjustments. Positive differences suggest recent price increases, while negative differences indicate reductions, reflecting possible discounting or competitive pricing strategies.

@@

```{r}
#| label: fig3
#| fig-cap: Each boxplot shows the distribution of numeric reliability grades for the top 5 pollsters, with individual poll grades represented by jittered points.
#| echo: false
#| warning: false
#| message: false

# Scatter plot of previous price vs. current price
# price_relationship_plot <- ggplot(analysis_data, aes(x = price_old, y = price_current)) +
#   geom_point(alpha = 0.4, color = "blue") +
#   geom_smooth(method = "lm", color = "red") +
#   labs(
#     title = "Relationship Between Previous and Current Prices",
#     x = "Previous Price (CAD)",
#     y = "Current Price (CAD)"
#   ) +
#   theme_minimal()
# price_relationship_plot

```
Figure 3: Relationship Between Previous and Current Prices. This scatter plot with a linear regression line shows the correlation between previous and current prices. A positive relationship would indicate that higher previous prices predict higher current prices, implying potential price stability or gradual price changes in the grocery market.

@@

```{r}
#| label: fig4
#| fig-cap: Each boxplot shows the distribution of numeric reliability grades for the top 5 pollsters, with individual poll grades represented by jittered points.
#| echo: false
#| warning: false
#| message: false

# Calculate mean and standard deviation of price_difference
# mean_diff <- mean(analysis_data$price_difference, na.rm = TRUE)
# sd_diff <- sd(analysis_data$price_difference, na.rm = TRUE)
# 
# # Filter out extreme values beyond 3 standard deviations
# filtered_data <- analysis_data %>%
#   filter(price_difference >= (mean_diff - 3 * sd_diff),
#          price_difference <= (mean_diff + 3 * sd_diff))
# 
# # Plot histogram with filtered data
# price_difference_hist <- ggplot(filtered_data, aes(x = price_difference)) +
#   geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
#   labs(
#     title = "Distribution of Price Differences (Filtered)",
#     x = "Price Difference (CAD)",
#     y = "Frequency"
#   ) +
#   theme_minimal()
# price_difference_hist

@@

Figure 4: Distribution of Price Differences (Filtered). This histogram illustrates the frequency of price changes across products, with extreme values beyond three standard deviations removed for clarity. The concentration around zero suggests that small price adjustments are most common, while the presence of both negative and positive values indicates instances of both price increases and decreases. This distribution provides insight into the stability and variability of pricing across vendors.
## Model Evaluation

To evaluate our model’s performance, we assess the predictive accuracy of the Bayesian linear regression model on the test data. The model’s Mean Absolute Error (MAE) and Root Mean Squared Error (RMSE) provide insights into the model’s ability to predict `price_current` based on `price_old`, `vendor_name`, and `price_difference`.

```{r}
#| label: fig5
#| fig-cap: Each boxplot shows the distribution of numeric reliability grades for the top 5 pollsters, with individual poll grades represented by jittered points.
#| echo: false
#| warning: false
#| message: false

#### Read Data ####
# Load only a sample of the data to reduce memory usage (e.g., 10% of the dataset)
# set.seed(853)
# 
# analysis_data <- fread("C:/Users/eliza/Downloads/hammer-3-compressed/cleaned_price_changes.csv")
# analysis_data <- analysis_data[sample(.N, size = 0.1 * .N)]
# 
# # Split data into training and testing sets
# train_indices <- sample(seq_len(nrow(analysis_data)), size = 0.8 * nrow(analysis_data))
# analysis_data_train <- analysis_data[train_indices]
# analysis_data_test <- analysis_data[-train_indices]
# 
# #### Model Data ####
# # Define a Bayesian regression model to predict `price_current` with fewer iterations and lighter priors
# 
# model <- stan_glm(
#   formula = price_current ~ price_old + as.factor(vendor_name) + price_difference,
#   data = analysis_data_train,
#   family = gaussian(),
#   prior = normal(location = 0, scale = 5, autoscale = TRUE),      # Slightly broader prior
#   prior_intercept = normal(location = 0, scale = 5, autoscale = TRUE),
#   prior_aux = exponential(rate = 0.5, autoscale = TRUE),          # Less informative prior
#   iter = 500,                                                     # Reduce iterations
#   chains = 2,                                                     # Fewer chains for faster computation
#   seed = 853
# )
# 
# #### Predict in Chunks ####
# # Define chunk size based on available memory
# chunk_size <- 20000  # Adjusted smaller chunk size for efficiency
# num_chunks <- ceiling(nrow(analysis_data_test) / chunk_size)
# 
# # Initialize a vector to store predictions
# predictions <- numeric(nrow(analysis_data_test))
# 
# # Loop through each chunk to generate predictions
# for (i in seq_len(num_chunks)) {
#   start <- ((i - 1) * chunk_size) + 1
#   end <- min(i * chunk_size, nrow(analysis_data_test))
#   
#   # Generate predictions for the current chunk
#   predictions[start:end] <- predict(model, newdata = analysis_data_test[start:end, ])
# }
# 
# #### Model Evaluation Metrics ####
# # Calculate Mean Absolute Error (MAE) and Root Mean Squared Error (RMSE)
# mae <- mean(abs(predictions - analysis_data_test$price_current))
# rmse <- sqrt(mean((predictions - analysis_data_test$price_current)^2))
# 
# # Store results in a data frame
# evaluation_metrics <- data.table(
#   Metric = c("Mean Absolute Error (MAE)", "Root Mean Squared Error (RMSE)"),
#   Value = c(round(mae, 2), round(rmse, 2))
# )
# 
# # Display the results table
# print(evaluation_metrics)

Table 1: Model Evaluation Metrics. The table displays the MAE and RMSE for the model, reflecting the average and squared prediction errors, respectively. Lower values indicate better predictive accuracy, suggesting that the model effectively captures the relationship between previous prices, vendor identities, and price differences.

These results demonstrate that vendor identity and previous prices are significant predictors of current prices in the Canadian grocery sector, with implications for understanding competitive strategies and consumer price sensitivity in the market.

@@
